---
title: "BMI of Dutch boys: solution"
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

<!--
%\VignetteEngine{rmarkdown::render}
%\VignetteIndexEntry{qgam_vignette}
-->
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
source(file = "my_settings.R")
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and plot it:
```{r dbbmi1, message = F}
library(mgcViz);
load(paste0(goodLocation, "/data/dbbmi.rda")) # change this

plot(bmi~age, data=dbbmi)
```

Fit basic gaussian GAM and plot predictions:
```{r dbbmi2, message = F}
fit1 <- gamV(bmi ~ s(age), data = dbbmi, aViz = list(nsim = 50))

plot(bmi~age, data=dbbmi)
lines(dbbmi$age, predict(fit1), col = 2)
```

2) Residuals checks
```{r dbbmi3, message = F}
qq(fit1)
```

Seems that residuals are skewed to the right. Let's do a conditional residual variance check:
```{r dbbmi3a, message = F}
check1D(fit1, "age") + l_gridCheck1D(gridFun = sd)
```
The variance of the residuals is increasing with age. Fit a model where we are modelling also the variance:
```{r dbbmi4, message = F}
fit2 <- gamV(list(bmi ~ s(age), ~ s(age)), data = dbbmi, family = gaulss, aViz = list(nsim = 50))
```

Repeat the check:
```{r dbbmi5, message = F}
qq(fit2)
check1D(fit2, "age") + l_gridCheck1D(gridFun = sd)
```

Non-constant variance problems seems more or less solved, but the residuals are still skewed to the right. 

3) Check whether we should increase bases dimension:
```{r dbbmi6, message = F}
tmp <- check(fit2)
```
It seems so. It seems that `bmi` increase sharply at is the first year of life, maybe we should adopt an adaptive smooth for the mean model? 
```{r dbbmi7, message = F}
fit3 <- gamV(list(bmi ~ s(age, k = 20, bs = "ad"), ~ s(age, k = 20)), 
             data = dbbmi, family = gaulss, aViz = list(nsim = 50))

AIC(fit2, fit3)
```
AIC improved and bases should be large enough now:
```{r dbbmi8, message = F}
tmp <- check(fit3)
```

Let's look at the fitted effects:
```{r dbbmi9, message = F}
print(plot(fit3), pages = 1)
```
Indeed, using an adaptive smooth made sense here.

4) Check skewness:
```{r dbbmi10, message = F}
library(e1071)
check1D(fit3, "age") + l_gridCheck1D(gridFun = skewness)
```

```{r dbbmi11, message = F}
library(mgcFam)
fit4 <- gamV(list(bmi ~ s(age, k = 20, bs = "ad"), ~ s(age, k = 20), ~ s(age), ~ 1), 
             data = dbbmi, family = shash, aViz = list(nsim = 50))

AIC(fit3, fit4)

qq(fit4)
```
Looks much better! And we get lower AIC. No re-plot smooth effects:
```{r dbbmi12, message = F}
print(plot(fit4), pages = 1)
```

Final check
```{r dbbmi12a, message = F}
check(fit4)
```
it all looks good.

5) Plotting quantiles:
```{r dbbmi13, message = F}
plot(bmi~age, data=dbbmi, col = "grey")
pr <- predict(fit4)
for(.q in c(0.01, 0.25, 0.5, 0.75, 0.99)){
  q_hat <- fit4$family$qf(.q, pr, wt = fit4$prior.weights, scale = 1)
  lines(dbbmi$age, q_hat, col = 2)
}
```