---
title: "Electricity demand modelling on GEFCom14 data: solution"
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

<!--
%\VignetteEngine{rmarkdown::render}
%\VignetteIndexEntry{qgam_vignette}
-->
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
source(file = "my_settings.R")
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and fit a basic Gaussian GAM:
```{r gefsmall1, message = F}
library(mgcViz)
load(paste0(goodLocation, "/data/gefcom_small.rda")) # change this

fit1 <- gamV(NetDemand ~ NetDemand.24 + Dow + Trend + s(wM) + s(wM_s95) + s(Posan, bs = "cc"), 
             data = gefcom_small, aViz = list(nsim = 50))

print(plot(fit1, allTerms = TRUE), pages = 1)
```

2) We check the mean residuals as a function of time (`trend`):
```{r gefsmall2, message = F}
check1D(fit1, "Trend") + l_gridCheck1D()
```

The long term trend seems to be non-linear. We also check whether we should increase the bases dimension:
```{r gefsmall3, message = F}
check(fit1)
```

`edf` is quite close to `k'` for all effects, and we get a very low p-value for `wM`. We have to increase `k` a bit.

3) Re-fit with a smooth long-term trend and increase basis dimension:
```{r gefsmall4, message = F}
fit2 <- gamV(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), data = gefcom_small, 
             aViz = list(nsim = 50))

AIC(fit1, fit2)

check1D(fit2, "Trend") + l_gridCheck1D()
```

AIC is much inproved, and there is no systematic residual pattern along `trend`. We recheck using `check`:
```{r gefsmall4a, message = F}
check(fit2)
```
Maybe we should increase the basis dimension for `wM` a bit more. But increasing the basis dimension for `Trend` too much is definitely not a good idea, because we would end up interpolating the data. 

4) Look at a qq-plot:
```{r gefsmall5, message = F}
qq(fit2)
```

We have some fairly fat tails, maybe better to adopt a scaled student-t distribution:
```{r gefsmall6, message = F}
fit3 <- gamV(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), data = gefcom_small, 
             family = scat,
             aViz = list(nsim = 50))

AIC(fit2, fit3)

qq(fit3)
```

Better AIC, and the QQ-plot looks much better. 

5) Worth considering the log-link, as an alternative to the identity link:
```{r gefsmall7, message = F}
fit4 <- gamV(NetDemand ~ NetDemand.24 + Dow + s(Trend, k = 6) + s(wM, k = 20) + 
             s(wM_s95, k = 15) + s(Posan, k = 15, bs = "cc"), 
             data = gefcom_small, 
             family = scat(link = log),
             aViz = list(nsim = 50))

AIC(fit3, fit4)
```
Seems better in terms of AIC. We finally plot the fitted effects:
```{r gefsmall8, message = F}
print(plot(fit4, allTerms = TRUE), pages = 1)
```