---
title: "Modelling rainfall in Parana state: solution"
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

<!--
%\VignetteEngine{rmarkdown::render}
%\VignetteIndexEntry{qgam_vignette}
-->
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
source(file = "my_settings.R")
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load packages and plot data
```{r bone1, message = F}
library(mgcViz)
load(paste0(goodLocation, "/data/parana.rda")) # change this

plot(parana[parana$TIME==1, ]$LO, parana[parana$TIME==1, ]$LA, xlab = "LO", ylab = "LA") 
```

We fit a median quantile GAM with an isotropic effect for longitude and latitude, a cyclic effect for the time of the year and smooth effects for distance from the sea and elevation:
```{r parana2, message = F}
fit1 <- gamV(PREC ~ s(LO, LA, k = 25) + s(seaDist) + s(TIME, bs = "cc") + s(EL), 
           data = parana)

print(plot(fit1), pages = 1)

qq(fit1)
```

```{r parana3, message = F}
fit2 <- gamV(list(PREC ~ s(LO, LA, k = 25) + s(seaDist) + s(TIME, bs = "cc") + s(EL),
                  ~ 1,
                  ~ 1),
           family = gevlss,
           data = parana)

print(plot(fit1), pages = 1)

qq(fit2, type = "tnorm")
```

Here using an isotropic smooth form `LO` and `LA` makes sense, because the two units are similar. We can still
check whether a tensor effect would lead to a different effect:
```{r parana3, message = F}
fit2 <- qgamV(PREC ~ te(LO, LA, k = c(5, 5)) + s(seaDist) + s(TIME, bs = "cc") + s(EL), 
                       data = parana, 
                       qu = 0.5)


library(gridExtra)
pl1 <- plot(sm(fit, 1)) + l_fitRaster() + l_fitContour() + theme(legend.position="bottom")
pl2 <- plot(sm(fit2, 1)) + l_fitRaster() + l_fitContour() + theme(legend.position="bottom")

grid.arrange(grobs = list(pl1$ggObj, pl2$ggObj), ncol = 2)
```
The two spatial effects are completely different! What happened is that the tensor product smooth got partially confounded with the distance from the sea effect. The `te` was more prone to doing this, because `seaDist` varies mostly along `LO` and it is mostly the marginal effect of `LO` that ended up offsetting the effect of `seaDist`.
In general if your model include a bivariate effect `s(x, y)` or `te(x, y)`, you have to be very careful when including an extra effect $s(f(x, y))$ where some $f$ is some fixed function (`seaDist` here).

We can of course fit this model to several quantiles:
```{r parana4, message = F}
fitM <- mqgamV(PREC ~ s(LO, LA, k = 25) + s(seaDist) + s(TIME, bs = "cc") + s(EL), 
               data = parana, 
               qu = seq(0.1, 0.9, length.out = 5), 
               err = 0.1)

plot(fitM, select = 1)
print(plot(fitM, select = 2:4), pages = 1)
```
Notice that the spatial effect seems much stronger for high quantiles than for the low one. The same is true for distance from the sea and seasonality (`TIME`), while the effect of elevation is not significant from `qu = 0.9`.
We can visualize the spatial effect in 3D as follows:
```{r parana4b, message = F}
plotRGL(sm(fitM[[5]], 1)) # This will not appear in the html file
```

We might also wonder whether the spatial effect changes with time. To verify this here we construct a tensor product smooth between the 2D thin-plate-spline spatial effect and the cyclical effect of time. We simplify the
model by removing the effect of `seaDist`.
```{r parana5, message = F}
fit4 <- qgamV(PREC ~ te(LO, LA, TIME, d = c(2, 1), k = c(20, 10), bs = c("tp", "cc")) + s(EL), 
              data = parana, 
              qu = 0.9, 
              err = 0.1)

plotSlice(sm(fit4, 1), 
          fix = list("TIME" = round(seq(1, 53, length.out = 6)))) + l_fitRaster()
```

You can plot any slice in 3D by doing:
```{r parana6, message = F}
plotRGL(sm(fit4, 1), fix = c("TIME" = 11)) # This will not appear in the html file
```

```

