---
title: "Retinopathy among diabetics: solution"
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

<!--
%\VignetteEngine{rmarkdown::render}
%\VignetteIndexEntry{qgam_vignette}
-->
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
source(file = "my_settings.R")
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and look at it:
```{r ret1, message = F}
library(mgcViz)
wesdr <- read.table(paste0(goodLocation, "/data/wesdr.txt")) # change this

pairs(wesdr)
```

2) Fit logistic model with all interactions
```{r ret2, message = F}
k = 7
fit1 <- gamV(ret ~ ti(dur,k=k) + ti(gly,k=k) + ti(bmi,k=k) + ti(dur,gly,k=k)+
                   ti(dur,bmi,k=k) + ti(gly,bmi,k=k), 
             data=wesdr,
             family=binomial(), 
             aGam = list(select=TRUE),
             aViz = list(nsim = 50))

summary(fit1)
```
Only last interaction seems important. Plot marginal effects:
```{r ret3, message = F}
print(plot(fit1, select = 1:3), pages = 1)
```

Plot interaction effects:
```{r ret4, message = F}
print(plot(fit1, select = 4:6), pages = 1)
```
First two interactions are essentially zero.

3) Fit simplified model and do residuals checks:
```{r ret5, message = F}
fit2 <- gamV(ret ~ s(dur,k=10) + te(gly,bmi,k=c(10, 10)), 
             data=wesdr,
             family=binomial(),
             aViz = list(nsim = 50))

pl <- list()
pl[[1]] <- ( check1D(fit2, "dur") + l_gridCheck1D(level = 0.95) )$ggObj      # OK
pl[[2]] <- ( check1D(fit2, "gly") + l_gridCheck1D(level = 0.95) )$ggObj      # OK
pl[[3]] <- ( check1D(fit2, "bmi") + l_gridCheck1D(level = 0.95) )$ggObj      # OK

library(gridExtra)
grid.arrange(grobs = pl, ncol = 2)
```
Looks relatively good. Checking across two variables:
```{r ret6, message = F}
check2D(fit2, "gly", "bmi") + l_gridCheck2D()
```
Big residuals in two locations... not sure it is worth increasing k.

QQ-plot with normally transformed residuals:
```{r ret7, message = F}
qq(fit1, type = "tnorm")

# shine(qq(fit1, type = "tnorm")) cannot do it in the html, but do try this
```

4) Plot tensor effect interactively:
```{r ret8, message = F}
plotRGL(sm(fit2, 2), residuals = TRUE)
rgl.close()
```
