---
title: "QGAM modelling of extreme rainfall in Switzerland: solution"
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

<!--
%\VignetteEngine{rmarkdown::render}
%\VignetteIndexEntry{qgam_vignette}
-->
  
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
```

```{r setup1, include=FALSE}
# Set workdir to source file location and then
source(file = "my_settings.R")
library(RhpcBLASctl); blas_set_num_threads(1)
```

1) Load data and fit a basic Gaussian GAM:
```{r uklss1, message = F}
library(mgcViz);
library(gamair);
data(swer)

form <- exra ~ s(nao) + s(elevation) + climate.region + s(E, N) + s(year, k = 5)
fit <- qgamV(form, data = swer, qu = 0.5)

summary(fit)
print(plot(fit), pages = 1)
```

2) Fit a smooth trend for each climate region:
```{r rus981}
fit2 <- qgamV(exra ~ s(nao) + s(elevation) +
                     s(year, climate.region, bs = "fs", k = 5) + s(E, N), 
                     data = swer, qu = 0.5)

summary(fit2)

AIC(fit) - AIC(fit2)

plot(sm(fit2, 3)) + l_fitLine(alpha = 1)
```
The effect of year-by-region is using relatively few degrees of freedom (`edf`), there might have been a slight decrease in rainfall in Valais, for instance.

3) Fit a 3D spatio-temporal tensor product effect:
```{r rus98221}
fit <- qgamV(exra ~ s(nao) + s(elevation) + climate.region +
                    te(E, N, year, d = c(2, 1), k = c(20, 5)), 
                    data = swer, qu = 0.5)

plotSlice(x = sm(fit, 3), 
          fix = list("year" = c(1985, 1995, 2005, 2015))) + l_fitRaster() + 
          l_fitContour() + l_points()
```
There doesn't seem to be much change in the precipication pattern, maybe there is a decrease of rainfall levels in the South East (in the Canton of Ticino).

4) We plot two slice through time using `rgl`:
```{r rus6346}
# These will not appear in the html output
plotRGL(x = sm(fit, 3), fix = c("year" = 1985), residuals = TRUE)
open3d()
plotRGL(x = sm(fit, 3), fix = c("year" = 2015), residuals = TRUE)
rgl.close()
rgl.close()
```
5) Remove the tensor effect w.r.t. time and fit the model to several quantiles:
```{r rus390}
fitM <- mqgamV(exra ~ s(nao) + s(elevation) + climate.region + s(E, N) + s(year, k = 5), 
               data = swer, qu = seq(0.1, 0.9, length.out = 9) )

# Plot univariate smooths
print(plot(fitM, select = c(1, 2, 4)), pages = 1)

summary(fitM[[9]])
```
It seems that the effect of `NAO` is very weak for extreme quantiles, however the reason for this might simply be that the data is quite sparse in the tails. Also, the effect of `year` seems to be highly non-linear for quantile 0.9, but `summary` shows that this effect is not significant. 

We can also plot the spatial effect:
```{r rus626}
print(plot(fitM, select = 3), pages = 1)
```
Interestingly the spatial effect is much stronger for high quantiles (extreme rainfall) than for the low ones. For quantile 0.9 the spatial effect varies between +50mm in the Canton of Ticino and -40mm in the Canton of Grisons. 

Finally we can see how the climate region effect changes depending on the quantile of interest:
```{r rus72547}
print(plot(fitM, select = 5), pages = 1)
```

